---
title: "Using Stars"
author: "Annie Kellner"
date: "6/22/2021"
output: html_document
---

[Helpful multi-page vignette](https://r-spatial.github.io/stars/articles/)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


```{r load libraries, echo=FALSE, message=FALSE, warning=FALSE}

rm(list = ls())

library(stars)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(viridis)

```

*NOTE: I recommend running this chunk by chunk, because some chunks of code will take a long time (several minutes) to run. Eval can also be set to false for time-consuming chunks. I considered storing everything as RData objects, but decided it was more instructive to keep the data raw. The only data stored as an R object is the park shapefile.*

## A little bit about stars objects

Stars objects can be created using a single file (e.g., one NetCDF) or multiple files (e.g., a folder of NetCDF's). The advantage to using the package stars is that metadata within the spatial objects can be manipulated using tidyverse syntax. Stars also accommodates complex data that may be (for example) multidimensional or curvilinear. 

This document uses data from NCAR from the WRST RSS to illustrate basic functions within the stars package. First, let's import a single NetCDF and take a look at the way data are structured in stars. Please note I have used the function `stars::read_stars()`to import this file. The function `stars::read_ncdf()` can also be used. The imported objects look identical, but there are some quirky differences in functionality. It is easy to convert one to the other, however, as I have done in the script called *summarize_multiple_years.R*. If you get errors in your code, it may be worth switching between these formats to see if the problem resolves. 

```{r import-file, warning=FALSE}

st <- read_stars("./data/met/ACCESS1-3_rcp45_BCSD_met_1950.nc4", sub = "tmax", curvilinear = c("longitude", "latitude"))
st
```

You can see that this data has 1 attribute (tmax) and three dimensions (x, y, and time). Attributes and dimensions are interchangeable and malleable as I will demonstrate later. Columns describing dimensions are as follows: 


**field**   |   **Meaning**
----------- |   -----------
from        |   origin index (e.g., 1)
to          |   final index (e.g., 365 for one year of daily data)
offset      |   start value for dimension (if regular; curvilinear x/y's are not a regular grid so will be 'NA')
delta       |   step size for dimension (if regular, as above)
refsys      |   geographic reference system
point       |   logical (T/F): whether cells refer to points or intervals. Is NA for this data (not sure why it's not FALSE)
values      |   sequence of values if grid is irregular (as NCAR data is)

  
## Subsetting data

Stars objects can be subsetted spatially using shapefiles or numerically using cell indices. I will proceed by spatially subsetting data using a park shapefile, but you can refer to *subset_stars_object.R* to see how to subset using cell indices. 

First, load the park shapefile. I saved this as a .Rds file so it could be easily accessed from GitHub for this tutorial. It was created using `sf::st_read()` and `dplyr::filter()`. Sf and stars play nicely together, as they were created by the same person. 

```{r}

wrst <- readRDS('./wrst.Rds') 

wrst <- st_shift_longitude(wrst) # script that changes longitudes in an sf object to 360 (rather than use negative numbers; this matches the NCAR data format)

plot(st_geometry(wrst)) # quick plot 

```


Make sure the shapefile lines up with the stars object appropriately:

```{r}

# Project if desired; I think everything still works if left as lat/long

wrst <- st_transform(wrst, 3338) # NAD83 Alaska Albers
st<- st_transform(st, 3338)

# Slice a single day from the stars object and overlay wrst to make sure everything lines up

day1 <- slice(st, along = time, 1)

plot(day1, border = NA, reset = FALSE) # reset = FALSE keeps the plot 'open' for more layers
plot(st_geometry(wrst), add = TRUE) 

```

Success! Subsetting is now very simple. The entire stars object can be subsetted using the same syntax. I am using the single slice here for illustrative purposes. The plot is small using the basic `plot()` function. 

```{r}

st_wrst <- day1[wrst] 
plot(st_wrst)

```

Visualization is improved using ggplot: 

```{r}

ggplot() + 
  geom_stars(data = st_wrst, alpha = 0.8) + 
  geom_sf(data = wrst, aes(), fill = NA) + 
  scale_fill_viridis() + 
  theme_map() +
  theme(legend.position = "bottom") +
  theme(legend.key.width = unit(2, "cm")) + 
  ggtitle("Tmax in WRST on January 1, 1950 (MIROC 5)")

```

The resolution is course because the underlying data are recorded in 12 km x 12 km cells.


## Combining multiple NetCDF files

It is essential to increase R's memory limit when you work with multiple NetCDF files. You can check how much memory R is allocated by calling the function `memory.limit()`. To increase the limit, use this code: `memory.limit(size = 60000)`. The memory limit will be reset when the R session is terminated. 60000 is an arbitrary number that worked on my machine. This number can be adjusted depending on the RAM/memory of the machine running the analysis.  

To easily load multiple relevant NetCDF files (e.g., all 'historical' files from a specific model/RCP), I find it easiest to manually move the files from one folder into another, from which you will load file names using the function `list.files()` After creating a vector of file names, you will read each .nc4 file in separately such that you create a list of n stars objects (n = number of years of data, or individual NetCDF files). 

```{r Read ncdf into list, message=FALSE, warning=FALSE}

file_list = list.files(path = "C:/Users/adillon/Documents/RSS/WRST/Test/met/Future", pattern = '.nc4', full.names = TRUE) # MIROC5 RCP4.5 years 2025 - 2055

l <- list() # Create a list to put the stars objects into

for(i in 1:length(file_list)){
  l[[i]] <- read_ncdf(file_list[i], var = "tmax", curvilinear = c("longitude", "latitude")) # need to read in as ncdf or coordinate system does not translate (not sure why)
}

```


Next, you will crop each stars object separately in order to create a similar list to the one above. After cropping, you need to convert each object using the `stars::st_as_stars()` function. This step is necessary to concatenate the list items into a single stars object. The single object enables 'raster math' (at a later step). 

```{r Crop and convert ncdf, message=FALSE, warning=FALSE}

cropped <- list() # create list for cropped stars objects

for(i in 1:length(l)){ # add cropped stars objects to a new list
  nc = l[[i]]
  nc = st_transform(nc, st_crs(wrst))
  nc_crop = nc[wrst]
  cropped[[i]] = nc_crop
}

cropped_st <- list()

for(i in 1:length(cropped)){
  cropped_st[[i]] <- st_as_stars(cropped[[i]])
}

```

Now you will combine each list item from `cropped` into a single stars object.

```{r}

rm(cropped, l, nc, nc_crop, day1, wrst, st, st_wrst) # remove large objects to free up memory

fut <- Reduce(c, cropped_st) # combine into single stars object 

fut

```

If you want to save the stars object for later use, you can do that using the code `write_stars(fut, "./data/fut_MIROC5")` (of course, the second argument is the file name, so this would change depending on personal organizational preference). The stars object can then be read in via `stars::read_stars()`. 


## Raster math

Summarizing data within a single stars object (e.g., fut) can be accomplished using `stars::st_apply()`. The third argument is the function desired, so can be any calculation in the same way `FUN` is specified in other R operations. Here, I am using `mean` to find the mean tmax value for each pixel. This can be conceptualized similarly to a rasterstack or brick, where each layer represents a time period (in this case, each layer = 1 day, for a total of 11322 layers) and the function is applied "vertically" through the layers. The result is then a stars object with a single layer representing the mean.

```{r}

fut_mean <- st_apply(fut, c("x", "y"), mean) # mean tmax value for each pixel. Second argument is the dimensions over which the function is applied. 

```


 To contrast future metrics with historical ones, you will need to follow all these instructions for the historical years (and model/rcp) of interest. Once you have two stars objects, you can easily calculate the change by simply subtracting the rasters from one another like so: `change <- fut_mean - hist_mean`. The end product will be a stars object with a single 'layer' that can be plotted in ggplot2 as above. 


## Summarizing data into dataframes for non-spatial analyses

Data can be easily retrieved from stars and analyzed via regular dataframes. To convert a stars object to a dataframe, you need only the `as.data.frame()` function. 

```{r}

wrst_df <- as.data.frame(st_wrst)
head(wrst_df)

```


From there, data can be aggregated using any standard methods. I am providing an example of aggregating data by time such that a single daily mean tmax is produced across all park pixels: 

```{r}

daily_mean_tmax <- wrst_df %>%
  na.omit() %>%
  group_by(time) %>%
  summarise(mean_tmax = mean(tmax))

head(daily_mean_tmax)

```


## Further notes

**On specifying variables to extract from NetCDF's**

I filtered the variable of interest (tmax, here) when creating the initial list of stars objects. If the variable was not specified, stars would create an object with as many attributes as variables contained in the data (e.g., tmax + tmin + precip = 3). I parceled tmax out in the interest of saving computing power when creating large stars objects, but if you are dealing with fewer years or substantial computing power, it's probably better to go through this process once and then `dplyr::select()` variables like so: `st %>% select(tmax) -> st_tmax`

**On manipulating variables**

Attributes and dimensions of stars objects are quite malleable. Attributes can be changed to dimensions and vice-versa, and also altered, added, or removed. Changing Celsius to Fahrenheit, for example, would simply require adding an attribute using tidy's `mutate()` function like so: `st %>% mutate(tmaxF = tmax * (9/5) + 32) -> stF`. Then this new attribute could be selected using `select()` as above. If converting between attributes and dimensions is required, please see *simple_summaries.R*. 




